The Key to a Data Parallel Compiler

### 2016-05-08 2016-spring-array/2016-array-key.tex
#1: Fixes identified by ispell.
#1 More fixes, typos, &c.

In the following sections we describe the core methods and idioms surrounding 
the AST encoding, construction of the Node Coordinate Matrix, and the use 
of these structures in handling nested, recursive AST relationships. This 
enables us to perform arbitrary computation over arbitrary sub-trees of an 
AST selected by parent-child relationships in the manner often seen in 
compiler transformations. These ideas have been further implemented and tested 
through the implementation of a complete data-parallel by construction 
compiler, called Co-dfns, that compiles a lexically scoped, functionally 
oriented dialect of APL with nested functions. 
\cite{hsu2014co,hsu2015accelerating}
The compiler targets both the
GPU and the CPU, and its core compiler is implemented in this pure, restricted 
language that uses only function composition over data-parallel primitives.

@@

That is, we indent the expression by 6 spaces, followed by the value
of the expression without indent. All expressions are evaluated right to 
left and all function application is infix; that is, a function 
application may apply a function (such as \verb;+;) to one or two arguments, 
called monadic and dyadic application respectively,
which must appear on the right and (optionally) on the left of the function 
name, as in the above example. Tables \ref{tab:scalarprims} to 
\ref{tab:conjunctions}
provide a listing of all of the primitives used in the Co-dfns compiler, 
a selection of which are used here.

@@

We introduce notation as needed throughout the exposition, so the reader 
is encouraged to refer to the tables as necessary to recall particular 
operations.

@@

The depth vectors for these trees we name \verb;Fd; and \verb;Ed;, respectively:

@@

The node types we call \verb;Ft; and \verb;Et;:

@@

And finally, we call the values vectors \verb;Fv; and \verb;Ev;:

@@

We combine these to form the respective AST matrices. We write \verb;A,B;
to catenate arrays \verb;A; and \verb;B; along their last axes and \verb;⍪A; to turn a
vector into a 1-column matrix. Thus, the two AST matrices are given
by the following expressions:

@@

The Depth Vector stores all edge information for the tree, but this information 
requires non-local access to utilize, such as traversing potentially the 
entire vector to determine the children of a specific node.
Node Coordinates fix this issue.

@@

\noindent
The above determines whether \verb;P; is an ancestor to \verb;C;.
The logical functions \verb;=;, \verb;∧;, and \verb;∨; are all extended 
point-wise over arrays.
The function \verb;⊢; returns its right argument.
The function \verb;(=∨0=⊢); reads as, ``equal or a zero right argument.'' 
We take this point-wise operation and reduce it with \verb;∧⌿; (For All).
This pattern is a special case of inner product, written \verb;f.g;. 
For example, \verb;+.×; is matrix multiplication. We transform our reduction and 
boolean function to a single predicate with the inner product operator:
\begin{verbatim}
      C∧.(=∨0=⊢)P
1
\end{verbatim}

@@

\begin{table*}
\centering
\begin{tabular}{l p{2.8in} l l l}
\toprule
Pass & Description & Core Traversal Pattern & Uses coordinates? \\
\midrule
Record Node Coordinates & Adds a new field to each node containing that node's 
 node coordinate. & Outer Product/Scan & Yes \\
Record Function Depths & Adds a new field to each node recording how many 
 functions surround the node. & Inner Product & Yes \\
Drop Unnamed Functions & Eliminates some code that will not be evaluated at 
 the top-level. & Filter & No \\
Drop Unreachable Code & Eliminates some unreachable code. &
 Filter/Inner Product & Yes \\
Lift Functions & Moves all functions to the top-level. & Key & Yes \\
Drop Redundant Nodes & Eliminates unnecessary nodes/nesting & Filter & No \\
Flatten Expressions & Removes nesting from expressions. & Key & Yes \\
Compress Atomic Nodes & Atomizes nested atom nodes & Amend & No \\
Propagate Constants & Inlines all references to literal values. &
 Amend/Rank & Yes \\
Fold Constants & Converts constant expressions to literals & Amend & No \\
Compress Expressions & Converts expression sub-trees into single nodes. & 
 Amend & No \\
Record Final Return Value & Records the value returned by each function. & 
 Key & No \\
Normalize Values Field & Normalizes the shape and size of the values field. &
 Amend & No \\
Lift Type-checking & Infers some type information at compile time &
 Power Limit/Rank & No \\
Allocate Value Slots & Does a form of frame allocation for variables &
 Key & No \\
Anchor Variables & Resolves lexically scoped variables &
 Key & Yes \\
Record Live Variables & Records the variables that are live at each point of 
 execution. & Key & No \\
Fuse Scalar Loops & Identifies Fusion opportunities and fuses expressions &
 Key & No \\
Type Specialization & Specializes each function for a series of potential inputs. 
 & Key & No \\
\end{tabular}
\caption{A listing of some compiler passes in the Co-dfns compiler and their 
 relationship with the Key operator and associated tree computation techniques}
\label{tab:passes}
\end{table*}

@@

To understand a bit better how the Key operator applies its function,
consider the function \verb;{⍺ ⍵}; which returns the pair of its right
and left arguments. If we apply it to the same value as above, we
get the following:

@@

In our case, we use either \verb;Fk; or \verb;Ek; as our keys applied to the
corresponding AST. We also will drop off the first row in each AST
using \verb;1↓; since this node ``contains'' everything. In a complete
AST this is usually the Module boundary node which contains the entire
set of functions and values in the module.

@@

The final task to complete function lifting of each function
body before lifting is to shift the depths in the depth vectors
to correspond to those of a function lifted to the top-level, and
to attach a function node to the top of each of the bodies. At that
point, we simply recombine all of the newly created functions into
a single top level.

@@

The compiler itself is composed of three main parts: the core compiler, 
the parser, and the code generator. The code generator includes the runtime 
library of the compiler. Parsing APL code in parallel is a well-studied 
problem. The compiler currently uses parsing combinators to do its parsing, 
though we plan to implement the Two-by-Two parser given in the literature. 
The code generator itself is a simple parallel map over each node, as 
each node may be generated independent from the other nodes, leading to a 
fully parallel generator, though there are some branching elements in the 
generator that still remain, currently.

@@

The current compiler represents a non-trivial instance of
the successful application of these techniques, but it focuses on an untyped, 
functionally oriented language. The type inference in the current version 
of the compiler is somewhat naive. More work is required to expand the 
type inferencer to include more sophisticated type inference. 
The compiler does not currently handle errors such as signals or 
exceptions, nor does it implement some more sophisticated compiler optimizations.
Further work remains to implement these techniques on an imperative and 
object-oriented language instead of a functional one. It would also be 
educational to implement a version for a lazy language.

@@

Bernecky \cite{bernecky2015abstract} argues that the increased programmability 
and desirable human factors of APL-style array programming can be implemented 
with competitive performance relative to more traditional techniques. This 
suggests that programs written in this highly abstract style may not suffer the 
performance gap traditionally assumed to go with their desirable features. 
This work is bolstered by previous work on the high-performance implemenation 
of array-oriented languages. 
\cite{ching1994experimental,ching1993primitive,
ching1990automatic,ju1991exploitation,
ju1991performance,bernecky1999reducing,schwarz1991acorn}

>>>>

In the following sections we describe the core methods and idioms surrounding 
the AST encoding, construction of the Node Coordinate Matrix, and the use 
of these structures in handling nested, recursive AST relationships. This 
enables us to perform arbitrary computation over arbitrary sub-trees of an 
AST selected by parent-child relationships in the manner often seen in 
compiler transformations. These ideas have been further implemented and tested 
through the implementation of a complete data-parallel by construction 
compiler, called Co-dfns, that compiles a lexically scoped, functionally 
oriented dialect of APL with nested functions 
\cite{hsu2014co,hsu2015accelerating}.
The compiler targets both the
GPU and the CPU, and its core is implemented in this pure, restricted 
language that uses only function composition over data-parallel primitives.

@@

\noindent
That is, we indent the expression by 6 spaces, followed by the value
of the expression without indent. All expressions are evaluated right to 
left and all function application is infix; that is, a function 
application may apply a function (such as \verb;+;) to one or two arguments, 
called monadic and dyadic application respectively,
which must appear on the right and (optionally) on the left of the function 
name, as in the above example. Tables \ref{tab:scalarprims} to 
\ref{tab:conjunctions}
provide a listing of all of the primitives used in the Co-dfns compiler, 
a selection of which are used here.

@@

\noindent
We introduce notation as needed throughout the exposition, so the reader 
is encouraged to refer to the tables as necessary to recall particular 
operations.

@@

\noindent
The depth vectors for these trees we name \verb;Fd; and \verb;Ed;, respectively:

@@

\noindent
The node types we call \verb;Ft; and \verb;Et;:

@@

\noindent
And finally, we call the values vectors \verb;Fv; and \verb;Ev;:

@@

\noindent
We combine these to form the respective AST matrices. We write \verb;A,B;
to catenate arrays \verb;A; and \verb;B; along their last axes and \verb;⍪A; to turn a
vector into a 1-column matrix. Thus, the two AST matrices are given
by the following expressions:

@@

\noindent
The depth vector stores all edge information for the tree, but this information 
requires non-local access to utilize, such as traversing potentially the 
entire vector to determine the children of a specific node.
Node Coordinates fix this issue.

@@

\noindent
The above determines whether \verb;P; is an ancestor to \verb;C;.
The logical functions \verb;=;, \verb;∧;, and \verb;∨; are all extended 
point-wise over arrays.
The function \verb;⊢; returns its right argument.
The function \verb;(=∨0=⊢); reads as, ``equal or a zero right argument.'' 
We take this point-wise operation and reduce it with \verb;∧⌿; (For All).
This pattern is a special case of inner product, written \verb;f.g;. 
For example, \verb;+.×; is matrix multiplication. We transform our reduction and 
Boolean function to a single predicate with the inner product operator:
\begin{verbatim}
      C∧.(=∨0=⊢)P
1
\end{verbatim}

@@

\begin{table*}
\centering
\begin{tabular}{l p{2.8in} l l l}
\toprule
Pass & Description & Core Traversal Pattern & Uses coordinates? \\
\midrule
Record Node Coordinates & Adds a new field to each node containing that node's 
 node coordinate. & Outer Product/Scan & Yes \\
Record Function Depths & Adds a new field to each node recording how many 
 functions surround the node. & Inner Product & Yes \\
Drop Unnamed Functions & Eliminates some code that will not be evaluated at 
 the top-level. & Filter & No \\
Drop Unreachable Code & Eliminates some unreachable code. &
 Filter/Inner Product & Yes \\
Lift Functions & Moves all functions to the top-level. & Key & Yes \\
Drop Redundant Nodes & Eliminates unnecessary nodes/nesting. & Filter & No \\
Flatten Expressions & Removes nesting from expressions. & Key & Yes \\
Compress Atomic Nodes & Atomizes nested atom nodes. & Amend & No \\
Propagate Constants & Inlines all references to literal values. &
 Amend/Rank & Yes \\
Fold Constants & Converts constant expressions to literals. & Amend & No \\
Compress Expressions & Converts expression sub-trees into single nodes. & 
 Amend & No \\
Record Final Return Value & Records the value returned by each function. & 
 Key & No \\
Normalize Values Field & Normalizes the shape and size of the values field. &
 Amend & No \\
Lift Type-checking & Infers some type information at compile time. &
 Power Limit/Rank & No \\
Allocate Value Slots & Does a form of frame allocation for variables. &
 Key & No \\
Anchor Variables & Resolves lexically scoped variables. &
 Key & Yes \\
Record Live Variables & Records the variables that are live at each point of 
 execution. & Key & No \\
Fuse Scalar Loops & Identifies Fusion opportunities and fuses expressions. &
 Key & No \\
Type Specialization & Specializes each function for a series of potential inputs. 
 & Key & No \\
\end{tabular}
\caption{A listing of some compiler passes in the Co-dfns compiler and their 
 relationship with the Key operator and associated tree computation techniques}
\label{tab:passes}
\end{table*}

@@

\noindent
To understand a bit better how the Key operator applies its function,
consider the function \verb;{⍺ ⍵}; which returns the pair of its right
and left arguments. If we apply it to the same value as above, we
get the following:

@@

\noindent
In our case, we use either \verb;Fk; or \verb;Ek; as our keys applied to the
corresponding AST. We also will drop off the first row in each AST
using \verb;1↓; since this node ``contains'' everything. In a complete
AST this is usually the Module boundary node which contains the entire
set of functions and values in the module.

@@

The final task to complete function lifting of each function
body before lifting is to shift the depths in the depth vectors
to correspond to those of a function lifted to the top-level and
to attach a function node to the top of each of the bodies. At that
point, we simply recombine all of the newly created functions into
a single top level.

@@

The compiler itself is composed of three main parts: the core compiler, 
the parser, and the code generator. The code generator includes the runtime 
library of the compiler. Parsing APL code in parallel is a well-studied 
problem. The compiler currently uses parsing combinators to do its parsing, 
though we plan to implement the Two-by-Two parser given in the literature
\cite{bunda1984apl}. 
The code generator itself is a simple parallel map over each node, as 
each node may be generated independent from the other nodes, leading to a 
fully parallel generator, though there are some branching elements in the 
generator that still remain, currently.

@@

The current compiler represents a non-trivial instance of
the successful application of these techniques, but it focuses on an untyped, 
functionally oriented language. The type inference in the current version 
of the compiler is somewhat na\"ive. More work is required to expand the 
type inferencer to include more sophisticated type inference. 
The compiler does not currently handle errors such as signals or 
exceptions, nor does it implement some more sophisticated compiler optimizations.
Further work remains to implement these techniques on an imperative and 
object-oriented language instead of a functional one. It would also be 
educational to implement a version for a lazy language.

@@

Bernecky \cite{bernecky2015abstract} argues that the increased programmability 
and desirable human factors of APL-style array programming can be implemented 
with competitive performance relative to more traditional techniques. This 
suggests that programs written in this highly abstract style may not suffer the 
performance gap traditionally assumed to go with their desirable features. 
This work is bolstered by previous work on the high-performance implementation 
of array-oriented languages. 
\cite{ching1994experimental,ching1993primitive,
ching1990automatic,ju1991exploitation,
ju1991performance,bernecky1999reducing,schwarz1991acorn}
